# Get Started {#chp:start}

::: epigraph
> Things are in the saddle,\
> And ride mankind.
>
> ---Ralph Waldo Emerson, *Ode, Inscribed to William H. Channing*
:::

R Markdown documents need to be compiled to Markdown before being rendered to a
target output format such as HTML and LaTeX.

For those who are familiar with R Markdown, you can think of the function
`litedown::fuse()` as the new `knitr::knit()`, `litedown::fiss()` as the new
`knitr::purl()`, and `litedown::mark()` as the new Pandoc. @fig:fuse-flow shows
the conversion process from R Markdown to a target format.

```{mermaid}
#| fuse-flow, echo = FALSE, fig.env = '.figure .box',
#| fig.cap = 'A diagram illustrating how **litedown** converts R Markdown documents to target output formats.'

flowchart LR
  A@{ shape: doc, label: "*.Rmd" }-. <code>fuse()</code> .-o F@{ shape: bolt }
  F o--- B@{ shape: doc, label: "*.md" }
  B-. <code>mark()</code> .-o G@{ shape: framed-circle }
  G --> C@{ shape: doc, label: "*.html" }
  G --> D@{ shape: doc, label: "*.tex" }
  G --> E@{ shape: doc, label: "..." }
  A-. <code>fiss()</code> .-o H@{ shape: cross-circ } --> I@{ shape: lin-doc, label: "*.R" }
```

The function `fuse()` "fuses" program code with narratives, i.e., it executes
all code in the source document and interweaves results with narratives in the
output document; `fiss()` splits the document and extracts the code. Similar to
**knitr**, **litedown** supports code chunks (@sec:code-chunks) and inline code
(@sec:inline-code).

We will first introduce `fuse()` in @chp:fuse, and then `mark()` in @chp:syntax
and @chp:mark.

::: callout-tip
If you do not do computing in Markdown, you do not need to learn
`fuse()`---`mark()` will be enough.
:::

## A new road to `roam()`

Before entering the **litedown** world, I suggest you forget the `Knit` button
if it has become your muscle memory to click on that button to view results
after writing every single line of code. The recommended way to work with
**litedown** is `litedown::roam()`, which allows you to preview and render
documents through a web interface.

To help you bid goodbye to the `Knit` button, please take 5 minutes to practice
the following mouse-clicking exercise (but try not to be addicted):

::: new-road
```{=html}
<p><button class="old-knit">Knit</button></p>
<p><button class="new-fuse">Render</button> <button class="new-fuse new-preview">Run</button></p>
```
:::

If the `Knit` button still sticks to your mind after you have practiced for a
few hundred times, I'd like to thank you for your loyalty to **knitr**, and
please see @sec:knit-button for how to continue to use the `Knit` button with
**litedown**.

You can learn more about the `Render` and `Run` buttons in @sec:live-preview.
Basically, the `Render` button renders a document to an output file, whereas the
`Run` button renders a document in memory without writing to disk.

::: callout-tip
Unless you use **litedown** programmatically (i.e., in scripts), you should
rarely need to call `fuse()` or `mark()` directly---`litedown::roam()` will call
`fuse()` and `mark()` for you behind the scenes.
:::

You can also create new documents with the [+](){.larger} button on the toolbar.

## A minimal example

To understand what `fuse()` does, we look at a minimal example:

`{r} .ex2(1)`

When we `fuse()` the source document, the program code will be executed, writing
results to the output.

You can see that the value `2` was assigned to the object `x` in a code chunk,
and the inline R expression `pi * x^2` was also evaluated to a number.

The advantage of interweaving computing code with text is that you can freely
update the source code according to the purpose of computing, and you will get
the up-to-date results after rebuilding the document. For example, you can
change `x` to `3`, and the area of the circle will be automatically updated
after you re-run `fuse()`. You do not need to copy and paste computational
outcome manually from one place to another.

## Basic concepts

A computational document contains both text narratives and code for computing.
@fig:flow-concepts shows the basic components of a document and how they work
together.

```{mermaid}
#| flow-concepts, echo = FALSE, fig.env = '.figure .box',
#| fig.cap = 'An illustration of the structure of a computational document and how its components are split, computed, and merged.'
flowchart TD
  O@{ shape: lin-doc, label: "Input document" }--"crack()"--> A@{ shape: f-circ } -.-o B[[YAML]]
  A -.-o C[[Text chunk]]
  C -.- C1>Text] --- C12{ } --> H[[Text output]]
  C -.- C2>Inline code]--> F{Value} --> H
  C -.- C3>More...] --- C32{ } --> H
  A -.-o D[[Code chunk]]
  D -.- D1>Chunk options]
  D -.- D2>Code]--> G{Results} --> I[[Chunk output]]
  A -.-o E[[More chunks...]] -.- K>...] --> L{...} --> M[[More output...]] -.-> J
  H -.-> J@{ shape: doc, label: "Markdown" }
  I -.-> J
  B -.-> P>...] --> Q{...} --> R[[YAML]] -.-> J
  J--"mark()" --> N@{ shape: tag-doc, label: "Output document" }
  subgraph "fuse()"
  C1
  C2
  C3
  D1
  D2
  K
  P
  end
  class O,J,N document
  classDef document stroke-width:3px
  class C12,F,C32,G,L,Q fuse-out
  classDef fuse-out fill:lightyellow
  class H,I,J,M,N,R output
  classDef output fill:none
```

First, the input is split into three types of blocks: the YAML metadata, text
chunks, and code chunks.

The *YAML metadata* contains information such as the title, author, date, and
other configurations for the document.

A *text chunk* could consist entirely of plain-text narratives that do not
require computing, or contain both text fragments and *inline code expressions*.
In the latter case, the inline code can be executed to return its value to the
output.

A *code chunk* contains code that often does more sophisticated computing than
inline code. Optionally, a code chunk can contain *chunk options* to control the
behavior of the computing and output.

After all code has been executed, the output blocks will be merged to a Markdown
document, which can be further rendered to a target output format such as HTML
and LaTeX.

:::: {#fig:humpty-dumpty .figure .box}
![Humpty
Dumpty](https://upload.wikimedia.org/wikipedia/commons/b/b2/Humpty_Dumpty_1_-_WW_Denslow_-_Project_Gutenberg_etext_18546.jpg)

::: caption
[ ](#@fig:humpty-dumpty) After raising two kids for a few years, I have been so
deeply brainwashed by nursery rhymes that it was Humpty Dumpty that came to my
mind when I was thinking about an example to explain computational documents.
Don't be surprised if you run across Blippi and garbage trucks somewhere in this
book.
:::
::::

Take the following document for example:

```` {.md data-file="*.Rmd"}
---
title: "A Report on the Wall"
author: "`{r} who`"
date: "`{r} Sys.Date()`"
---

```{r, order = 0}
who = 'Humpty Dumpty'
```

`{r} who` sat on a wall.

`{r} who` had a great fall.

```{r}
broken = strsplit(who, ' ') |> unlist()
```

All the king's horses and all the king's men

Couldn't put `{r} broken[1]` together again.

<!--
If the king's horses and men knew R, they could've put
Humpty together again by paste(broken, collapse = ' ').
-->
````

The YAML metadata is provided between `---` and `---` in the beginning. Note
that it can also contain code to be evaluated, such as the variable `who` and
the function call `Sys.Date()`.

The document has two code chunks in triple-backtick (```` ``` ````) fences. The
first code chunk contains a chunk option `order = 0` (you may see
@sec:option-order for its meaning later). You can do any computing in code
chunks and generate text results or graphics.

There are two text chunks: one between the two code chunks, and one after the
second code chunk. These text chunks contain inline R code expressions in
`` `{r} ` ``. An inline code expression is typically used to get a single value.

In the end, there is a comment in `<!-- -->`. Comments will not be displayed in
the output document.

## A few words on naming

When I started writing the **litedown** package, I was thinking of nuclear
fusion reactions (although my knowledge in physics is quite limited): text and
code chunks are like atomic nuclei, and they can be combined to form a report.
The process is controlled by a reactor. That was how I came up with the function
names `fuse()` and `reactor()`.

To prepare the atomic nuclei, I need to `crack()` the source document. When the
input is not a document but a script, I will `sieve()` out the elements from
comments. The fusion reaction can take place after `crack()` or `sieve()` is
done.

Maybe the neurons in my brain are wired in an odd way, but I feel mysteriously
happy that the names `crack()` and `sieve()` weakly resembles my friend Carson
Sievert's name. For that reason, Carson will live in my heart forever, even if
he would never send a pull request to **litedown**.

Since there also exist fission reactions, the function `fiss()` is meant to
decompose a document.

As for the package name "litedown", it has three meanings:

1.  The obvious meaning is the lightweight of the package.

2.  The less obvious meaning is that light comes down from the fusion reactor
    like the Sun. This meaning is represented by the symbol for `fuse()`, which
    is a zigzag arrow (↯).

3.  Finally, it is a play on the phrase "lie down". I intended to write this
    package as my last "down" package. After it is done, I wish to lie down.

Let there be light.
