# CSS/JS add-ons {#chp-addon}

The **litedown** package aims at lightweight with a minimal number of features
at its core, but you can add more features by yourself. In this chapter, we
introduce some CSS/JS add-ons from the repository
<https://github.com/yihui/misc.js>. You can load arbitrary external JS and CSS
files via the `js` and `css` meta variables. There are numerous JS libraries and
CSS frameworks on the web that you can use, and you do not have to use the ones
mentioned in this chapter. You can also write CSS/JS by yourself to enhance your
HTML applications.

Remember that the CSS and JS are introduced under the output format
`litedown::html_format` in YAML metadata, e.g.,

``` yaml
---
output:
  litedown::html_format:
    meta:
      css: ["one.css", "two.css"]
      js: ["three.js", "four.js"]
---
```

For the sake of brevity, we will omit the full YAML fields in examples
throughout this chapter but only use the `css` and `js` fields. A file name
`foo.js` denotes the file under the `js/` directory of the aforementioned
`yihui/misc.js` repository. Similarly, `foo.css` is under the `css/` directory.

## HTML slides

With `snap.css` and `snap.js`, you can create lightweight HTML slides:

``` yaml
css: ["@default", "@snap"]
js: ["@snap"]
```

You can learn more in `vignette('slides', package = 'litedown')`.

## HTML articles

We can style an HTML page in an article format via the following CSS and JS:

``` yaml
css: ["@default", "@article"]
js: ["@sidenotes", "@appendix"]
```

The `article.css` is mainly for styling the article frontmatter, body, and side
content.

-   The `sidenotes.js` is required only if you want to place certain elements on
    the left or right side, such as the table of contents (TOC), footnotes, and
    sidenotes.

-   The `appendix.js` is required only if you have an appendix in the article.

The web version of this book is also based on the article format, so you know
what an article format looks like when you read the HTML version of the book.

### The overall style

The maximum width of the article body is 800px. For larger screens, this means
there will be extra space in the left/right margin, where we can place auxiliary
information, such as the TOC and footnotes. On smaller screens, the side content
will be collapsed into the body.

The article frontmatter, body, and optionally the appendix are placed in
separate boxes.

The default typeface is sans-serif, and you can customize it by supplying an
external CSS file (via the `css` meta variable) or just embedding CSS in the
document body, e.g.,

```` md
```{css, echo=FALSE}
body {
  font-family: Palatino, "Book Antiqua", Georgia, serif;
  font-size: 1em;
}
```
````

### Side elements

The TOC and footnotes are automatically placed in the margin if space permits.
You can also write arbitrary content in the margin via a fenced `Div`.

#### The TOC

The TOC is sticky on the left side as you scroll down the article. If you do not
like this behavior, you may cancel it via CSS:

``` css
#TOC {
  top: unset;
}
```

#### Footnotes

Footnotes are moved to the right side. When you move your cursor over a footnote
number in the body, the footnote will be moved next to your cursor. This can be
convenient when you have multiple footnotes on a line, since you do not need to
look for a specific footnote in the margin.

#### Arbitrary sidenotes

You can write anything in the margin by using a fenced `Div` with the classes
`.side` and `.side-left` or `.side-right`.

::: {.side .side-left}
**Notice**

Here is a note on the left side. Anything permitted by law is permitted here.
Math? No problem!

$$e^{i\theta}=\sin{\theta}+i\cos{\theta}$$

When you have this sidenote "hammer", I'm sure you will hit a lot of nails into
the margin, even if you do not have to.
:::

::: {.side .side-right}
And here is a note on the right side. Seriously, we should let
[commonmark](https://commonmark.org)'s authors know that fenced `Div`s really
deserve first-class support! They can make Markdown infinitely customizable.
:::

``` markdown
::: {.side .side-left}
**Anything** on the left.
:::
```

``` markdown
::: {.side .side-right}
_Anything_ on the right.
:::
```

### Body elements

Inside the article body, you can write a few special elements.

#### Full-width elements

When an element is wider than the article body, you can show it in its full
width by enclosing the element in a fenced `Div` with the class `.fullwidth`,
e.g.,

``` markdown
::: {.fullwidth}
![text](path/to/image)
:::
```

::: fullwidth
![Sunspots](https://prose.yihui.org/post/2020/11/10/r-markdown-demo/index_files/figure-html/sunspots-1.svg)
:::

If you use R Markdown, you can generate a wide plot or table from an R code
chunk, e.g.,

```` markdown
::: {.fullwidth}
```{r}
#| sunspots, echo=FALSE, fig.dim=c(14, 4),
#| fig.cap='Monthly mean relative sunspot numbers from 1749 to 1983.'
par(mar = c(4, 4, .1, .1), bg = 'lightgoldenrodyellow', fg = 'red', las = 1)
plot(sunspots, col = 'red', panel.first = grid())
```
:::
````

If you want to show code (`echo = TRUE`) but do not want the code to be in the
full-width container, you can apply the `.fullwidth` class to the plot only,
e.g.,

```` md
```{r}
#| sunspots, fig.dim=c(14, 4), fig.env='.fullwidth .figure .box',
#| fig.cap='Monthly mean relative sunspot numbers from 1749 to 1983.'
par(mar = c(4, 4, .1, .1), bg = 'lightgoldenrodyellow', fg = 'red', las = 1)
plot(sunspots, col = 'red', panel.first = grid())
```
````

#### Left/right quotes

::: quote-right
> Whenever you find that you are on the side of the majority, it is time to
> pause and reflect.
>
> ---[Mark Twain](http://www.quotationspage.com/quotes/Mark_Twain/81)
:::

Sometimes you may want to add a quote but do not want it to take the full width
in the body. You may use a fenced `Div` with the class `.quote-left` or
`.quote-right`.

Despite the class names, the content does not have to be a quote. If you do want
a quote, just use the blockquote syntax `>`, e.g.,

``` markdown
::: {.quote-right}
> This is a boring quote.
>
> ---Someone
:::
```

#### Margin embedding

::: embed-right
```{r, echo=FALSE}
head(mtcars, 4)
```
:::

You can embed elements on the left or right margin using a fenced `Div` with the
class `.embed-left` or `.embed-right`. These elements will float to the left or
right and exceed the margin by about 200px, which can save some space in the
article body. You can use the extra space to explain the embedded element with
text.

We have embedded a table of the first 4 rows of the `mtcars` data on the right
margin, which you can see if the browser window is wide enough.

## Tabbed sections

You can load the script `tabsets.js` and CSS `tabsets.css` to create tabsets
from sections (see [documentation
here](https://yihui.org/en/2023/10/section-tabsets/)).

``` yaml
css: ["@tabsets"]
js: ["@tabsets"]
```

## Code folding

Code folding is supported by `fold-details.js` (see [documentation
here](https://yihui.org/en/2023/09/code-folding/)).

``` yaml
js: ["@fold-details"]
```

## Callout blocks

A callout block is a fenced Div with the class name `callout-*`. Callouts
require `callout.css` and `callout.js`:

``` yaml
css: ["@callout"]
js: ["@callout"]
```

For example:

``` md
::: callout-tip
This is a tip.

> You can write arbitrary content, such as a blockquote.

::: callout-caution
Even another callout!
:::

Is that cool?
:::
```

Output:

::: callout-tip
This is a tip.

> You can write arbitrary content, such as a blockquote.

::: callout-caution
Even another callout!
:::

Is that cool?
:::

The stylesheet `callout.css` supports styling three types of callouts: `tip`,
`caution`, and `important`. For example:

::: callout-caution
Be careful when testing for strict equality of floating point numbers.

```{r}
seq(0, 1, .2) == c(0, .2, .4, .6, .8, 1)
```
:::

::: callout-important
For the sake of reproducibility, please remember to render an R Markdown
document in a *new R session* before publishing or submitting the output
document.
:::

You do not have to use `callout.css` but can define your own CSS rules, e.g.,

``` css
.callout-important {
  background-color: red;
  color: yellow;
}
```

Under the hood, `callout.js` turns the fenced Div into a `<fieldset>` for the
form:

``` html
<fieldset class="callout-*">
  <legend>Title</legend>

  Content.
</fieldset>
```

The content comes from the original fenced Div. The title comes from the class
name (converted to uppercase) by default. You can provide a custom title via the
`data-legend` attribute of the Div, e.g.,

``` md
::: {.callout-tip data-legend="Information"}
:::
```

The icons before the callout titles can be defined via CSS, e.g., you can add
two exclamation marks before the title of `important` callouts:

``` css
.callout-important legend::before {
  content: "!! ";
}
```

The default icons defined in `callout.css` are essentially [UTF-8
characters](https://www.w3schools.com/charsets/ref_utf_symbols.asp). In theory,
there are hundreds of thousands of characters that you can choose from. Each
character is 1 to 4 bytes. For example, you can define a `music` callout with
the music note symbol ♫ in the CSS:

``` css
.callout-music {
  background-color: springgreen;
}
.callout-music legend::before {
  content: "♫ ";
}
```

Then you can insert a `music` callout in your document:

``` md
::: callout-music
Please listen to this lovely song.
:::
```

## Right-align a quote footer

You can use the script `right-quote.js` to right-align a blockquote footer if it
starts with an em-dash (`---`).

``` yaml
js: ["@right-quote"]
```

## Add anchor links to headings

The CSS is necessary only if you want to hide the anchors by default and reveal
them on hover.

``` yaml
css: ["@heading-anchor"]
js: ["@heading-anchor"]
```

## Style keyboard shortcuts

The script `key-button.js` identifies keys and the CSS styles them, which can be
useful for [showing keyboard
shortcuts](https://yihui.org/en/2023/02/key-buttons/).

``` yaml
css: ["@key-buttons"]
js: ["@key-buttons"]
```

Of course, you can combine any number of JS scripts and CSS files if you want
multiple features.
