# CSS/JS assets {#chp:assets}

::: epigraph
> And it seemed as though in a little while the solution would be found, and
> then a new and splendid life would begin; and it was clear to both of them
> that they had still a long, long road before them, and that the most
> complicated and difficult part of it was only just beginning.
>
> ---Anton Chekhov, *The Lady with the Dog*
:::

The **litedown** package aims at lightweight with a minimal number of features
at its core, but you are free to add more features by yourself. In this chapter,
we introduce some CSS/JS assets from the GitHub repository
<https://github.com/yihui/lite.js>. You can load arbitrary external JS and CSS
files via the `js` and `css` meta variables. There are numerous JS libraries and
CSS frameworks on the web, and you do not have to use the ones mentioned in this
chapter. You can also write CSS/JS by yourself to enrich your HTML applications.

Remember that the CSS and JS are introduced under the output format `html` in
YAML metadata, e.g.,

``` yaml
---
output:
  html:
    meta:
      css: ["one.css", "two.css"]
      js: ["three.js", "four.js"]
---
```

For the sake of brevity, we will omit the full YAML fields in examples
throughout this chapter but only use the `css` and `js` fields. A file name
`foo.js` denotes the file under the `js/` directory of the aforementioned
[`lite.js` repository](https://github.com/yihui/lite.js). Similarly, `foo.css`
is under the `css/` directory.

All these CSS/JS resources can be used offline, and there are several ways to do
it. One way is to clone the GitHub repo to your working directory, and use the
files you need, e.g.,

``` yaml
js: ["repo/path/js/callout.js"]
```

You may also see @sec:offline and @sec:embed-resources for other ways.

## A quick reference {#sec:asset-reference}

@tab:assets provides a quick reference of existing CSS/JS assets in the
`lite.js` repository. See @sec:meta-css for the meaning of the `@` character in
`css` / `js` names.

```{r}
#| assets, echo = FALSE, print.args = I(list(limit = Inf)),
#| tab.env = '.table .no-ref-label',
#| tab.cap = 'A reference of CSS/JS assets in the `lite.js` repository.'
a = litedown:::assets
for (i in c('css', 'js')) a[, i] = ifelse(is.na(a[, i]), NA, paste0('`', a[, i], '`'))
a = cbind(feature = rownames(a), a, section = paste0(' @sec:', rownames(a)))
rownames(a) = NULL
a
```

If the source document is R Markdown instead of Markdown, an alternative way (in
addition to configuring YAML metadata) to add assets is to call the function
`litedown::vest()` in a code chunk. You can either pass vectors of `css` / `js`
asset names to the `css` / `js` arguments, or the `feature` names in @tab:assets
to the `feature` argument, e.g.,

```` md
```{r, echo = FALSE}
# add assets by feature
litedown::vest(feature = c('callout', 'copy-button'))
# or add css/js directly
litedown::vest(css = c('@callout', '@book'), js = '@callout')
```
````

You can call `vest()` multiple times in a document to keep adding assets.

## The default CSS {#sec:default}

The default stylesheet defines the basic styles of the document body, lists,
code, figures, tables, block quotes, TOC, links, and cross-references, etc. For
example, the maximum width for the body is set to 800px. If you want to override
it, you can provide your own stylesheet, e.g.,

``` css
body {
  max-width: 1000px;
}
```

The default CSS is included when no `css` is provided in the YAML metadata. You
can use either `@default` or `default` to apply the default CSS.

``` yaml
css: ["@default"]
# or
css: ["default"]
```

The former uses the jsdelivr resource and will not be embedded in HTML by
default. The latter points to the `default.css` shipped with the **litedown**
package, and will always be embedded.

## Copy buttons {#sec:copy-button}

Normally you would have to select the text before copying it, and copy buttons
can save the effort of text selection when you want to copy the text from an
HTML element. You just click on the button, and the text is automatically
copied. Copy buttons can be added via CSS/JS:

``` yaml
css: ["@copy-button"]
js: ["@copy-button"]
```

By default, this will add copy buttons to all code blocks (`<code>` in `<pre>`)
and all elements with the class name `copy-this`. For example, below are two
quotes, and the right quote is inside a fenced Div `::: copy-this`, so you will
see a copy button in the right quote if you hover over it with the cursor:

:::: flex-col
> Love like you've never been hurt\
> Dance like nobody's watching\
> Sing like nobody's listening\
> Work like you don't need the money\
> Live like it's heaven on earth
>
> ---Alfred D'Souza

::: copy-this
> Copy like you've never been hurt\
> Copy like nobody's watching\
> Copy like nobody's listening\
> Copy like you don't need the money\
> Copy like it's heaven on earth
>
> ---Yihui (Ode to the [Copy
> Ninja](https://en.wikipedia.org/wiki/Kakashi_Hatake))
:::
::::

All code blocks in this book also have copy buttons in them.

In fact, you can add the copy button to any element on an HTML page. You may
read [this post](https://yihui.org/en/2023/09/copy-button/) if you are
interested in generalizing the copy button to other elements (without using the
`copy-this` class).

## Tabsets {#sec:tabsets}

A tabset provides a compact way to arrange content in tabs, which can save a lot
of vertical space on the page. To create a tabset, you need to load the
following assets:

``` yaml
css: ["@tabsets"]
js: ["@tabsets"]
```

There are two ways to create a tabset in Markdown.

### From a bullet list

The first way is to write a bullet inside a fenced Div with the class name
`tabset`. The first element of each bullet item will become the tab title, and
the rest of elements will become the tab pane, e.g.,

```{md}
::: tabset
- Tab one

  Content: -OOOO0OOO⍥
  
  Another paragraph.

- Tab two <!--active-->

  Content: -OOOOOOOO⍥
  
  - A normal bullet list.
:::
```

By default, the first tab is the active tab initially. To specify a different
initial active tab, add a comment `<!--active-->` to the bullet item.

Note that all content within the respective tabset needs to be indented,
including any code blocks, e.g.,:

````{md}
::: tabset
- Iris

  ```{r}
  plot(iris)
  ```
:::
````

You can use the `Left` (Left) and `Right` (Right) Arrows on your keyboard to
navigate through the tabs when the tabset is on focus (e.g., after you have
clicked on it once). Now keep pressing the Left or Right Arrow key, and you will
realize that there is a caterpillar crawling in the example tabset above (and
also looking at you). It's cute, isn't it?

Since bullet lists can be nested, you can also nest one tabset in another. To do
that, you just wrap a sub-list in a `tabset` div, e.g.,

``` md
::: tabset
- Tab one

  ::: tabset
  - Child tab one
  
  - Child tab two
  :::

- Tab two
:::
```

I'm not sure why anyone would want to create nested tabsets, but it sounds like
a great way to create a maze. Perhaps you can write a detective story or game
with it?

Wait a minute. I just came up with an idea! We can use nested tabsets to
navigate through a recursive list. I have implemented this idea in the function
`xfun::tabset()`, and felt for the first time of my life that I dare to peruse
the huge `recordPlot()` list now! That said, I don't want to scare you with
`recordPlot()`, but let's take a look at the config file of this book as an
example instead:

```{r}
cfg = xfun::taml_file('_litedown.yml')
xfun::tabset(cfg, dput)
```

I have never loved YAML, but this time, I feel that YAML is adorable.

### From sections

The second way to create a tabset is through section headings. If you add a
class name `tabset` to a heading, headings of the next (lower) level will be
used as the tab titles, e.g.,

``` md
## Demo tabs {.tabset}

### Tab one

Content.

### Tab two {.active}

Content.
```

If a heading has the `active` class, it will be the active tab initially.

You can also create a tabset by providing an empty fenced Div with the class
name `tabset`, and subsequent headings will be converted to tabs, e.g.,

``` md
::: tabset
:::

## Tab one

Content.

## Tab two

Content.
```

The tabset will end before it hits a higher-level heading. The tabset in the
first example above can be ended by a `##` heading, and the second example can
be ended by a `#` heading.

To explicitly end a tabset without using a higher-level heading, you can insert
a comment of the form `<!--tabset:ID-->` to the end of the tabset, where `ID` is
the ID of the starting element of the tabset, e.g.,

``` md
## Demo tabs {.tabset #foo}

### Tab one

### Tab two

<!--tabset:foo-->

### Not a tab
```

``` md
::: {.tabset #foo}
:::

## Tab one

## Tab two

<!--tabset:foo-->

## Not a tab
```

When a tabset is created from sections, the tab titles (from section headings)
will appear in the TOC of the document if TOC is enabled (@sec:toc).

### Code generated

Tabsets are especially useful when presenting multiple images or tables.
In these cases, it might be desired to generate tabsets programatically.

This can be achieved by using the `results = asis` and generating the required format:

````{md}
::: tabset
```{r, results = "asis", echo = "FALSE"}
for(letter in letters){
  cat("*", letter, "\n\n")
  cat("  ", letter, "\n")
  cat("  ", "or other content\n\n")
}
```
:::
````

Using these technique, almost any content bar figures can be generated
and embedded within tabsets.

Figures are automatically placed at the end of code block, which means they would be placed
after the tabset.
To circumvent this, you need to pregenerate images either manually (using `png()`, `pdf()` or other device) using named code chunk, and then link them by constructing link with
the known name.

````{md}
```{r, mtcars_hist, include = FALSE}
# Generate images
Map(
    f = hist,
    mtcars,
    xlab = names(mtcars),
    main = sprintf("Histogram of \"%s\"", names(mtcars))
)
```

```{r, echo = FALSE, results = "asis"}
for(i in seq_along(mtcars)){
  cat(sprintf("* %s (%d)\n", names(mtcars)[i], i))
  fig_name = paste0(reactor("fig.path"), "mtcars_hist-", i, ".png")
  cat(sprintf("  ![](%s)\n\n", fig_name))
}
```
````

## Code folding {#sec:fold-details}

``` yaml
js: ["@fold-details"]
```

You can fold any element with the JS. See [this
post](https://yihui.org/en/2023/09/code-folding/) for how to configure the
script.

## Callout blocks {#sec:callout}

A callout block is a fenced Div with the class name `callout-*`. Callouts
require `callout.css` and `callout.js`:

``` yaml
css: ["@callout"]
js: ["@callout"]
```

For example:

```{md, attr.asis = '.callout-output'}
::: callout-tip
This is a tip.

> You can write arbitrary content, such as a blockquote.

You can even _nest another callout_ into this one!

Is that cool?
:::
```

### Built-in callouts

The stylesheet `callout.css` supports styling `tip`, `note`, `caution`,
`warning`, `important`, and `example` callouts. For example:

::: callout-note
Thank you for your notice! Your notice has been noted.
:::

::: callout-caution
Be careful when testing for strict equality of floating point numbers.

```{r}
seq(0, 1, .2) == c(0, .2, .4, .6, .8, 1)
```
:::

::: callout-warning
Never try to out-stubborn a cat!
:::

::: callout-important
For the sake of reproducibility, please remember to render an R Markdown
document in a *new R session* before publishing or submitting the output
document.
:::

::: callout-example
```{md}
::: {.callout-example data-legend="Demo"}
Change the title of this example to "Demo".
:::
```
:::

### Customizing callouts

You do not have to use `callout.css` but can define your own CSS rules, e.g.,

``` css
.callout-important {
  background: red;
  color: yellow;
}
```

Under the hood, `callout.js` turns the fenced Div into a `<fieldset>` for the
form:

``` html
<fieldset class="callout-*">
  <legend>Title</legend>

  Content.
</fieldset>
```

The content comes from the original fenced Div. The title comes from the class
name (converted to uppercase) by default. You can provide a custom title via the
`data-legend` attribute of the Div, e.g.,

``` md
::: {.callout-tip data-legend="Information"}
:::
```

The icons before the callout titles can be defined via CSS, e.g., you can add
two exclamation marks before the title of `important` callouts:

``` css
.callout-important legend::before {
  content: "!! ";
}
```

The default icons defined in `callout.css` are essentially [UTF-8
characters](https://www.w3schools.com/charsets/ref_utf_symbols.asp). In theory,
there are hundreds of thousands of characters that you can choose from. Each
character is 1 to 4 bytes. For example, you can define a `music` callout with
the music note symbol ♫ in the CSS:

``` css
.callout-music {
  background: springgreen;
  border-color: yellow;
}
.callout-music legend::before {
  content: "♫ ";
}
```

Then you can insert a `music` callout in your document:

``` md
::: callout-music
Please listen to this lovely song.
:::
```

Alternatively, if you are using `callout.css`, you can also use CSS variables to
define the border color, background, and icon of a callout, e.g.,

``` css
.callout-music {
  --callout-background: springgreen;
  --callout-border: yellow;
  --callout-icon: "♫ ";
}
```

## Add anchor links to headings {#sec:heading-anchor}

The CSS is necessary only if you want to hide the anchors by default and reveal
them on hover.

``` yaml
css: ["@heading-anchor"]
js: ["@heading-anchor"]
```

## HTML articles {#sec:article}

You can style an HTML page in an article format via the following CSS and JS:

``` yaml
css: ["@default", "@article"]
js: ["@sidenotes", "@appendix"]
```

The `article.css` is mainly for styling the article frontmatter, body, and side
content.

-   The `sidenotes.js` is required only if you want to place certain elements on
    the left or right side, such as the table of contents (TOC), footnotes, and
    sidenotes.

-   The `appendix.js` is required only if you have an appendix in the article.

The web version of this book is also based on the article format, so you know
what an article format looks like when you read the HTML version of the book.

### The overall style

The maximum width of the article body is 800px. For larger screens, this means
there will be extra space in the left/right margin, where we can place auxiliary
information, such as the TOC and footnotes. On smaller screens, the side content
will be collapsed into the body.

The article frontmatter, body, and optionally the appendix are placed in
separate boxes.

The default typeface is sans-serif, and you can customize it by supplying an
external CSS file (via the `css` meta variable) or just embedding CSS in the
document body, e.g.,

```` md
```{css, echo = FALSE}
body {
  font-family: Palatino, "Book Antiqua", Georgia, serif;
  font-size: 1em;
}
```
````

### Side elements

The TOC and footnotes are automatically placed in the margin if space permits.
You can also write arbitrary content in the margin via a fenced `Div`.

#### The TOC

The TOC is sticky on the left side as you scroll down the article. If you do not
like this behavior, you may cancel it via CSS:

``` css
#TOC {
  top: unset;
}
```

#### Footnotes on the side

Footnotes are moved to the right side. When you move your cursor over a footnote
number in the body, the footnote will be moved next to your cursor. This can be
convenient when you have multiple footnotes on a line, since you do not need to
look for a specific footnote in the margin.

#### Arbitrary sidenotes

You can write anything in the margin by using a fenced `Div` with the classes
`.side` and `.side-left` or `.side-right`.

::: {.side .side-left}
**Notice**

Here is a note on the left side. Anything permitted by law is permitted here.
Math? No problem!

$$e^{i\theta}=\sin{\theta}+i\cos{\theta}$$

When you have this sidenote "hammer", I'm sure you will hit a lot of nails into
the margin, even if you do not have to.
:::

``` markdown
::: {.side .side-left}
**Anything** on the left.
:::
```

``` markdown
::: {.side .side-right}
_Anything_ on the right.
:::
```

In case you are interested in the technical detail, it's quite simple to move an
element into the margin using CSS. For example, the `.side-right` class above is
roughly defined as:

``` css
.side-right {
  width: 200px;
  float: right;
  margin-right: -200px
}
```

That basically means the width of the element is 200px and it floats to the
right. Now its right side will touch the right margin of its parent element (the
article body). What we need to do next is move it further to the right by 200px
(i.e., its width), which is done by the `-200px` right margin. Remember, a
positive right margin in CSS moves an element to the left, and a negative right
margin moves it to the right.

### Body elements

Inside the article body, you can write a few special elements.

#### Full-width elements

When an element is wider than the article body, you can show it in its full
width by enclosing the element in a fenced `Div` with the class `.fullwidth`,
e.g.,

``` markdown
::: {.fullwidth}
![text](path/to/image)
:::
```

::: fullwidth
![Sunspots](https://prose.yihui.org/post/2020/11/10/r-markdown-demo/index_files/figure-html/sunspots-1.svg)
:::

If you use R Markdown, you can generate a wide plot or table from an R code
chunk, e.g.,

```` markdown
::: {.fullwidth}
```{r}
#| sunspots, echo = FALSE, fig.dim = c(14, 4),
#| fig.cap = 'Monthly mean relative sunspot numbers from 1749 to 1983.'
par(mar = c(4, 4, .1, .1), bg = 'lightyellow', fg = 'red', las = 1)
plot(sunspots, col = 'red')
grid()
```
:::
````

If you want to show code (`echo = TRUE`) but do not want the code to be in the
full-width container, you can apply the `.fullwidth` class to the plot only,
e.g.,

```` md
```{r}
#| sunspots, fig.dim = c(14, 4), fig.env = '.fullwidth .figure',
#| fig.cap = 'Monthly mean relative sunspot numbers from 1749 to 1983.'
par(mar = c(4, 4, .1, .1), bg = 'lightyellow', fg = 'red', las = 1)
plot(sunspots, col = 'red')
grid()
```
````

#### Left/right quotes

::: quote-right
> Whenever you find that you are on the side of the majority, it is time to
> pause and reflect.
>
> ---[Mark Twain](http://www.quotationspage.com/quotes/Mark_Twain/81)
:::

Sometimes you may want to add a quote but do not want it to take the full width
in the body. You may use a fenced `Div` with the class `.quote-left` or
`.quote-right`.

Despite the class names, the content does not have to be a quote. If you do want
a quote, just use the blockquote syntax `>`, e.g.,

``` markdown
::: {.quote-right}
> This is a boring quote.
>
> ---Someone
:::
```

#### Margin embedding

You can embed elements on the left or right margin using a fenced `Div` with the
class `.embed-left` or `.embed-right`. These elements will float to the left or
right and exceed the margin by about 200px, which can save some space in the
article body. You can use the extra space to explain the embedded element with
text.

::: embed-right
```{r, echo = FALSE}
mtcars[1:4, 1:8]
```
:::

We have embedded a table of the first 4 rows of the `mtcars` data on the right
margin, which you can see if the browser window is wide enough.

## HTML books {#sec:book}

``` yaml
css: ["@book"]
```

## Chapter TOC {#sec:chapter-toc}

``` yaml
js: ["@chapter-toc"]
```

## Highlight TOC items {#sec:toc-highlight}

``` yaml
js: ["@toc-highlight"]
```

## Paged HTML {#sec:pages}

``` yaml
css: ["@pages"]
js: ["@pages"]
```

## HTML slides {#sec:snap}

With `snap.css` and `snap.js`, you can create lightweight HTML slides:

``` yaml
css: ["@default", "@snap"]
js: ["@snap"]
```

You can learn more in `vignette('slides', package = 'litedown')`.

## Right-align a quote footer {#sec:right-quote}

You can use the script `right-quote.js` to right-align a blockquote footer if it
starts with an em-dash (`---`).

``` yaml
js: ["@right-quote"]
```

> Without a little madness, life is not worth living. Let us follow the guidance
> of our inner voice. Why should we fry each of our actions like a piece of cake
> on a sensible frying pan?
>
> ---Milan Kundera, *Immortal*

## Center images {#sec:center-img}

``` yaml
js: ["@center-img"]
```

## Open external links in new windows {#sec:external-link}

``` yaml
js: ["@external-link"]
```

## Style keyboard shortcuts {#sec:key-buttons}

The script `key-button.js` identifies keys and the CSS styles them, which can be
useful for [showing keyboard
shortcuts](https://yihui.org/en/2023/02/key-buttons/).

``` yaml
css: ["@key-buttons"]
js: ["@key-buttons"]
```

Of course, you can combine any number of JS scripts and CSS files if you want
multiple features.

## Include arbitrary CSS/JS {#sec:arbitrary-assets}

You can include arbitrary CSS frameworks and JS libraries in your documents, not
limited to aforementioned ones in the GitHub repository `yihui/lite.js`. Below
is a quick example of using the
[simple-datatables](https://github.com/fiduswriter/simple-datatables/) library
to create interactive tables:

```` md
---
title: Display Tables with the JS Library DataTables
output:
  html:
    meta:
      css: ["@default", "@npm/simple-datatables/dist/style"]
      js: ["@npm/simple-datatables"]
---

::: {#mtcars-table}
```{r}
I(mtcars)
```
:::

```{js, type = "module"}
new simpleDatatables.DataTable('#mtcars-table > table');
```
````

See @sec:simple-dt for more ways to create this type of table.
