# Books and Websites {#chp:sites}

Books and websites are usually based on multiple input files under a directory.
For a directory to be recognized as a book or website project, it needs to
contain a configuration file named `_litedown.yml`.

If you want to customize the output formats `html_format` or `latex_format` for
books or websites, you should do it in `_litedown.yml`, e.g.,

``` yaml
output:
  litedown::html_format:
    options:
      toc:
        depth: 4
  litedown::latex_format:
    meta:
      documentclass: "book"
```

## Books

The `_litedown.yml` file should contain a top-level field named `book`, which
currently supports these options:

``` yaml
book:
  new_session: false
  subdir: false
  pattern: "[.]R?md$"
  chapter_before: "Information before a chapter."
  chapter_after: "This chapter was generated from `$input$`."
```

You can choose whether to render each input file in a new R session, whether to
search subdirectories for input files, the types of input files (e.g., you can
use `.md` or `.R` files if you want), and additional information to be included
before/after each chapter, in which you can use some variables such as
`$input$`, which is the path of each input file.

### Previewing a single chapter

Rendering a whole book may be time-consuming[^06-site-1] and unnecessary when
you work on a book. It may be easier to only preview the single chapter that you
currently work on. The preview can be done with `litedown::roam()`:

[^06-site-1]: Remember that you can cache time-consuming code chunks to make the
    rendering faster.

-   If you run the index file (e.g., `index.Rmd`), the whole book will be
    previewed.

-   If you run other chapter files, only the specific chapter is previewed.

When a chapter contains dependencies on certain elements in other chapters
(e.g., a chapter includes cross-references to other chapters), we recommend that
you run the index file to preview the whole book at least once before you
preview individual chapters, to make **litedown** know the book elements fully.

Similarly, if you use the configuration `new_session: false` to render all
chapters in the same R session, and a later chapter uses computed results from a
previous chapter, you will need to preview the full book at least once to make
sure the results are computed and exist in the R session.

You may also choose to run the index file to preview the whole book but then
work on an individual chapter. In this case, `roam()` will try to detect changes
in chapter files. When a certain chapter file has been updated, its output on
the full book page will be updated. That is, the full book page is *partially*
updated without being reloaded. This method lets you preview changes in one
chapter while presenting the whole book. However, please note that certain JS
libraries may not work well in this preview mode. When in doubt, refresh the
page.

### R package documentation

R package developers can build the full package documentation as a book. A
series of helper functions have been provided in **litedown** to get various
information about the package, such as the package description (`pkg_desc()`),
news (`pkg_news()`), citation (`pkg_citation()`), source code (`pkg_code()`),
and all manual pages (`pkg_manual()`). You can call these functions in code
chunks to print out the desired information. For example, you may put them in
the appendix (@sec:appendices):

```` md
# Appendix {.appendix}

# Package Metadata

```{r, echo=FALSE}
litedown::pkg_desc()
```

To cite the package:

```{r, echo=FALSE}
litedown::pkg_citation()
```

# News

```{r, echo=FALSE}
litedown::pkg_news(recent = 0)  # show full news
```

# Manual pages

```{r, echo=FALSE}
litedown::pkg_manual()
```

# Source code

```{r, echo=FALSE}
litedown::pkg_code()
```
````

## Websites

The `_litedown.yml` file should contain a top-level field named `site`, and you
are likely to customize the `meta` variables `css`, `include_before`, and
`include_after` for the `html_format`, e.g.,

``` yaml
site:
  rebuild: "outdated"
  pattern: "[.]R?md$"

output:
  litedown::html_format:
    meta:
      css: ["@default"]
      include_before: "[Home](/) [About](/about.html)"
      include_after: "&copy; 2024 | [Edit]($input$)"
```

Basically, `include_before` can take a file or text input that will be used as
the header of each web page, and `include_after` will be the footer.
