# Books and Websites {#chp:sites}

::: epigraph
> I can at least listen without indignation to the critic who is of the opinion
> that when one surveys the aims of cultural endeavour and the means it employs,
> one is bound to come to the conclusion that the whole effort is not worth the
> trouble, and that the outcome of it can only be a state of affairs which the
> individual will be unable to tolerate.
>
> ---Sigmund Freud, *Civilization and Its Discontents*
:::

Books and websites are usually based on multiple input files under a directory.
For a directory to be recognized as a book or website project, it needs to
contain a configuration file named `_litedown.yml`.

If you want to customize the output formats `html_format` or `latex_format` for
books or websites, you should do it in `_litedown.yml`, e.g.,

``` yaml
output:
  html:
    options:
      toc:
        depth: 4
  latex:
    meta:
      documentclass: "book"
```

## Books

The `_litedown.yml` file should contain a top-level field named `book`, which
currently supports these options:

``` yaml
book:
  new_session: false
  subdir: false
  pattern: "[.]R?md$"
  chapter_before: "Information before a chapter."
  chapter_after: "This chapter was generated from `$input$`."
```

You can choose whether to render each input file in a new R session, whether to
search subdirectories for input files, the types of input files (e.g., you can
use `.md` or `.R` files if you want), and additional information to be included
before/after each chapter, in which you can use some variables such as
`$input$`, which is the path of each input file.

### Previewing a single chapter

Rendering a whole book may be time-consuming[^07-site-1] and unnecessary when
you work on a book. It may be easier to only preview the single chapter that you
currently work on. The preview can be done with `litedown::roam()`:

[^07-site-1]: Remember that you can cache time-consuming code chunks to make the
    rendering faster.

-   If you run the index file (e.g., `index.Rmd`), the whole book will be
    previewed.

-   If you run other chapter files, only the specific chapter is previewed.

When a chapter contains dependencies on certain elements in other chapters
(e.g., a chapter includes cross-references to other chapters), we recommend that
you run the index file to preview the whole book at least once before you
preview individual chapters, to make **litedown** know the book elements fully.

Similarly, if you use the configuration `new_session: false` to render all
chapters in the same R session, and a later chapter uses computed results from a
previous chapter, you will need to preview the full book at least once to make
sure the results are computed and exist in the R session.

You may also choose to run the index file to preview the whole book but then
work on an individual chapter. In this case, `roam()` will try to detect changes
in chapter files. When a certain chapter file has been updated, its output on
the full book page will be updated. That is, the full book page is *partially*
updated without being reloaded. This method lets you preview changes in one
chapter while presenting the whole book. However, please note that certain JS
libraries may not work well in this preview mode. When in doubt, refresh the
page.

## Websites

The `_litedown.yml` file should contain a top-level field named `site`, and you
are likely to customize the `meta` variables `css`, `include_before`, and
`include_after` for the `html_format`, e.g.,

``` yaml
site:
  rebuild: "outdated"
  pattern: "[.]R?md$"

output:
  html:
    meta:
      css: ["@default"]
      include_before: "[Home](/) [About](/about.html)"
      include_after: "&copy; 2024 | [Edit]($input$)"
```

Basically, `include_before` can take a file or text input that will be used as
the header of each web page, and `include_after` will be the footer.

## R package documentation {#sec:pkg-site}

R package developers can build the full package documentation as either a book
or a website.

### The `pkg_*()` helper functions

A series of helper functions have been provided in **litedown** to get various
information about the package, such as the package description (`pkg_desc()`),
news (`pkg_news()`), citation (`pkg_citation()`), source code (`pkg_code()`),
and all manual pages (`pkg_manual()`). You can call these functions in code
chunks to print out the desired information.

For example, you may call them in the appendix (@sec:appendices) of a book:

```` md
# Appendix {.appendix}

# Package Metadata

```{r, echo = FALSE}
litedown::pkg_desc()
```

To cite the package:

```{r, echo = FALSE}
litedown::pkg_citation()
```

# News

```{r, echo = FALSE}
litedown::pkg_news(recent = 0)  # show full news
```

# Manual pages

```{r, echo = FALSE}
litedown::pkg_manual()
```

# Source code

```{r, echo = FALSE}
litedown::pkg_code()
```
````

Alternatively, you can build a package website by calling these functions in
separate `.Rmd` files, e.g., `pkg_desc()` in `index.Rmd`, `pkg_news()` in
`news.Rmd`, and so on.

### Package sites via GitHub Action

You may use GitHub Actions to automatically build and deploy package websites.
The key is to call `litedown::fuse_site()`. I have provided [a simple GitHub
action](https://github.com/yihui/litedown/blob/main/site/action.yml) in the
**litedown** repository, and you can add a step `uses: yihui/litedown/site@HEAD`
to your GitHub workflow to use it. Below is [the full
workflow](https://github.com/yihui/litedown/blob/main/.github/workflows/github-pages.yml)
for building and publishing **litedown**'s site to GitHub Pages:

```{embed, attr.output = '.yaml data-file="github-pages.yml" data-source="1"'}
"../.github/workflows/github-pages.yml"
```

You can copy it to the `.github/workflows/` directory of the R package's Git
repository and commit/push it via Git. Within a few minutes, the package site
should be ready at `https://user.github.io/pkg/`, where `user` is your GitHub
username, and `pkg` is the repository name, e.g.,
<https://yihui.github.io/litedown/>.

The action `yihui/litedown/site` supports a few options that you can configure
under the `with` field. In the above example workflow, we used the option
`site-dir`. Below are the options currently supported:

```{r, echo = FALSE, results = 'asis'}
inputs = xfun::yaml_load(readLines('../site/action.yml'), use_yaml = FALSE)$inputs
cat(sprintf(
  '- `%s`: %s (default: `%s`).', names(inputs),
  unlist(lapply(inputs, `[[`, 'description')),
  unlist(lapply(inputs, `[[`, 'default'))
), sep = '\n\n')
```

By default, the site is built under the `site/` directory of the package. If you
change this path, please remember to adjust the `path` option for the
`actions/upload-pages-artifact` action accordingly.

If the site source directory in your package does not exist or is empty, the
action will copy [the site
template](https://github.com/yihui/litedown/tree/main/site) from **litedown**,
which contains the following `.Rmd` files (explained in @tab:pkg-site):

```{r}
list.files('../site/', '[.]Rmd$')
```

:::: {.table .box}
::: caption
[ ](#@tab:pkg-site) Explanations of site source files.
:::

| source | description | content |
|----|----|----|
| `index` | home page | `pkg_desc()`, `pkg_citation()`, and `README.md` |
| `articles` | package vignettes | the `doc/` folder of the installed package |
| `code` | source code (R, C, C++, etc.) | `pkg_code()` |
| `manual` | help pages | `pkg_manual()` |
| `news` | news | `pkg_news()` |
| `_footer` | page footer | package authors |
::::

You can freely customize this template according to your own needs (e.g.,
modify, add, or delete files), and add the folder to your repository. Then the
action will use your custom template to build the site.

The `exclude` option allows you to exclude certain pages from the site. By
default, `code.Rmd` is excluded. You can exclude more pages. For example, if
your package does not have vignettes, you can specify
`exclude: 'code.Rmd articles.Rmd'` (if you do not want `code.Rmd`, either).

The `cleanup` option allows you to run a command before publishing the site,
e.g., you can delete the `.Rmd` source and `.yml` config files.
