---
title: "litedown: R Markdown Reimagined"
author: "Yihui Xie"
date: "`{r} Sys.Date()`"
---

```{r, include = FALSE}
if (packageVersion('xfun') < '0.49.7') stop(
  'package versions too low; please install newer versions:\n',
  "install.packages(c('litedown', 'xfun'), repos = 'https://yihui.r-universe.dev')"
)

options(width = 80)

# return an example file in fenced Div
.ex = function(n, ext = '.Rmd', last = NA) {
  f = Sys.glob(sprintf('../examples/%03d*', n))
  f = f[endsWith(f, ext)]
  if (length(f) != 1) stop('cannot find the example ', n)
  x = xfun::read_utf8(f)
  i = which(x == '<!-- ... -->')[1]
  if (ext == '.Rmd') {
    if (!is.na(i)) x = x[-(i + 0:1)]
  } else {
    partial = TRUE
    if (!is.na(last)) x = tail(x, last) else {
      if (!is.na(i)) x = x[-(1 + 0:i)] else partial = FALSE
    }
    if (partial) x = c(rep('.', 3), x)
  }
  x = litedown:::trim_blank(paste(x, collapse = '\n'))
  xfun::fenced_block(x, c(
    sub('^([.])R(.+)', '\\1\\2', ext), sprintf('data-file="%03d%s"', n, ext)
  ))
}
# source and output in the same div
.ex2 = function(n, ext = c('.Rmd', '.md'), ...) {
  x1 = .ex(n, ext[1])
  x2 = .ex(n, ext[2], ...)
  xfun::fenced_div(c(x1, x2), '.flex-col')
}
```

# Preface {.unnumbered}

::: callout-important
The **litedown** package is still new and experimental. The documentation is
very incomplete. Besides, **litedown** was designed for minimalists with a
limited scope. Average users should perhaps consider using **rmarkdown** or
Quarto instead.
:::

> You may say I'm a dreamer\
> But I'm not the only one\
> I hope someday you'll join us\
> And the world will live as one
>
> ---John Lennon, *Imagine*

Imagine there's no PDF. It's easy if you try. No Word below us. Above us, only
HTML.

I do not mean PDF and Word are bad. I only lament the time that human beings
spend on various document formats, given how much a static web page can do.

## The past journey

Having worked on the development of R Markdown for nearly 12 years, I have to
confess that I have become a little perplexed about my work, although I clearly
remember the exciting moments. For example:

-   The moment when I added syntax highlighting to R code in `.Rnw` documents.
    Beautiful!

-   The moment when I managed to generate a base R plot and a ggplot in the same
    code chunk and place them side by side in LaTeX. Magic! Not to mention that
    ggplot does not require an explicit `print()` call in a code chunk---you
    probably do not even know what that means now.

-   The moment when I saw the impossible-to-read diagram that shows all possible
    conversions among numerous document formats on the Pandoc homepage.
    Incredible!

-   The moment when I discovered DZSlides in Pandoc. Wow! So PowerPoint and
    LaTeX beamer were not the only choices for slides.

-   The moment when some rstudio::conf attendees started applauding after I told
    them that the PowerPoint support had been added to the development version
    of **rmarkdown** when one person asked about it.

-   The moment when I came across remark.js. And tufte.css. And GitBook. And
    Hugo. And distill.pub. And **htmlwidgets**, **reticulate**, and so on.

-   ...

I think the work was largely meaningful. The only problem is I do not see an
end. The list of exciting things to do goes on and on.

## A question

"At forty, I had no more doubts." Confucius said. Apparently, I'm not Confucius.
On the contrary, as I'm turning forty, I'm having more doubts. I have been
reflecting on the things that have kept me busy.

If I were to summarize the 700+ episodes of Naruto in one sentence, I would use
the question that Gaara (the Fifth Kazekage) asked Onoki (the Third Tsuchikage):

> When did you all forsake yourselves?

In my opinion, this question was the most critical turning point in the 700+
episodes, reminding the 79-year-old Onoki and the rest of people of their
original dreams. The shinobi world existed to put an end to wars, but it turned
out to bring even more and larger-scale wars.

When did we forsake simplicity? We create software to simplify things, but
software often ends up in feature wars.

Markdown was originally invented for simplicity. That is, to make it easier to
write HTML. I used to take out my wallet and tell people that if they are unable
to learn the basics of Markdown in 5 minutes, I'd award them 20 dollars. I also
used to say "In HTML I Trust", which sounds like a joke, but I really love HTML
and web technologies. Again, I do not mean other document formats are bad at
all. It is just that I feel HTML has the greatest potential, and I hope to take
full advantage of its power.[^1]

[^1]: At one night, as I was thinking about Pandoc Lua filters, an obvious fact
    suddenly came to my mind: suppose all I care about is HTML output, then the
    good old JavaScript can actually play a perfect role of Lua filters, because
    you can manipulate the DOM arbitrarily with JavaScript.

In other words, I do not expect that "the (software) world will live as one". I
just want to make the HTML world a little bit better.

## Overview

The **litedown** package is an attempt to reimagine R Markdown with one primary
goal—do HTML, and do it well, with the minimalism principle. Before LaTeX fans
walk away in disappointment, let me quickly clarify that LaTeX output is also
supported, but please do not expect anything fancy before you learn to customize
LaTeX templates. Most other output formats are not supported. No Word, RTF,
PowerPoint, or EPUB.

R Markdown is rendered via `litedown::fuse()`, which is similar to
`rmarkdown::render()` and `knitr::knit()`. Markdown is rendered via
`litedown::mark()`, which uses **commonmark** instead of Pandoc.

The **commonmark** package follows the GFM (GitHub Flavored Markdown) spec,
which can be seen as a subset of Pandoc's Markdown. Therefore the **litedown**
package can be viewed as a small subset of the existing R Markdown ecosystem
(the latter is based on Pandoc). It aims at simplicity, lightweight, and speed,
at the cost of giving up some advanced features. This package is intended for
minimalists. Most users may prefer using tools based on Pandoc instead, such as
**rmarkdown** or Quarto, which offer richer features.

Is **litedown** really simple? From the developer's perspective, yes, it is,
largely due to the limited scope of the package. From the user's perspective,
some features are definitely not that simple. However, the point is that the
core is simple and small, and you can enable or disable most features. What's
more, you can implement features by yourself if you know CSS/JS.

## Scope

You can view **litedown** as a minimal (re-)implementation of some core packages
in the existing R Markdown ecosystem, such as **rmarkdown** for reports,
**xaringan** for slides, **bookdown** for books, **blogdown** for websites,
**pkgdown** for R package sites, and **pagedown** for paged HTML documents.

$$\mathrm{litedown} = \min{\{R\}} + \{D_i\} - \{D_e\} + \{J\}$$

-   $R$ = **knitr** + **evaluate** + **rmarkdown** + **bookdown** +
    **blogdown** + **pagedown** + **pkgdown** + **xaringan** + **tufte** +
    **distill**

-   $D_i$ = (internal dependencies) **commonmark** + **xfun**

-   $D_e$ = (external dependencies) Pandoc + Bootstrap + jQuery + GitBook +
    Hugo + paged.js + remark.js + tufte.css + distill.js/.css + ...

-   $J$ = Lightweight [vanilla JavaScript/CSS](https://github.com/yihui/lite.js)

It is absolutely not the goal for **litedown** to become a substitute of tools
based on **knitr** and Pandoc, such as **rmarkdown** and Quarto. If you are not
sure if you should choose **litedown** or **rmarkdown**/Quarto, you may want to
choose the latter (especially Quarto).

### Features out of scope

Output formats besides HTML and LaTeX are unlikely to be supported.[^2] If other
output formats are desired, you may use Pandoc to do the conversion.

[^2]: In fact, `xml,` `man`, `text`, and `commonmark` output formats are
    supported (thanks to the **commonmark** package), but perhaps they are not
    very useful to average users.

For tables, only pipe tables are supported. Other table formats are not
recognized.

At the moment, **litedown** mainly supports R as the computing language. Other
languages might be added in the future, but please keep your expectation low,
because the support is unlikely to be as good as R. Back in 2013, I experimented
with a simple idea to run program code in other languages, and the outcome was
the [**runr**](https://github.com/yihui/runr) package. However, I forsook
simplicity after other sophisticated packages emerged, such as **reticulate**
and **JuliaCall**. I may revisit the idea in the future.

HTML widgets are not supported yet, but may be reimagined in the future with
some minimal support. Meanwhile, you are free to add arbitrary JS libraries to
your documents (@sec:arbitrary-assets), so it is entirely possible that you can
create HTML widgets on your own.

## Highlights

### Small footprint

Almost everything in **litedown** was written from scratch. The package is very
lightweight, with only two R package dependencies: **commonmark** and **xfun**.

It is a deliberate design choice to keep this package lightweight, to make it
relatively easy to use and simple to maintain. The functions `mark()` and
`fuse()` can be viewed as significantly trimmed-down versions of Pandoc and
**knitr**, respectively.

Just to give you a better idea about the "lightweight" (the numbers below refer
to *uncompressed* software and source code as of today):

-   As the Markdown converter, Pandoc 3.5 is about 180Mb, and **commonmark** is
    about 1Mb on my macOS.

-   Bootstrap's CSS is about 12,000 lines (plus 1500 lines of JS), and
    **litedown**'s base CSS is about 100 lines (no JS).

-   For paginating HTML documents,
    [paged.js](https://cdn.jsdelivr.net/npm/pagedjs/) has 33,000+ lines of code,
    and my naive implementation,
    [pages.js](https://github.com/yihui/lite.js/blob/main/js/pages.js), has
    about 400 lines (200 lines of JS and 200 lines of CSS).

Admittedly, pursuing lightweight requires sacrifice in features and quality
(e.g., my pages.js is far less sophisticated than paged.js), but overall I feel
the trade-off should be reasonable to those who prefer minimalism. With a 2Mb
footprint (the total file size of **litedown**, **commonmark**, and **xfun**),
you can:

-   Write basic components of Pandoc's Markdown (@chp:syntax), including
    headings, table of contents (TOC), lists, tables, pipe figures, code blocks,
    quotes, fenced Divs, LaTeX math, footnotes, and citations, etc.

-   Automatically number and cross-reference arbitrary elements
    (@sec:cross-references), including but not limited to headings, figures,
    tables, and equations.

-   Perform computing with code chunks and inline code expressions (@chp:fuse).

-   Create single-file reports/slides, or multiple-file books/websites
    (@chp:sites).

If you load additional JS/CSS assets, you can create more types of elements,
such as folded code blocks, tabsets, side elements, sticky TOC, and callout
blocks (@chp:assets).

### Live preview everything

To get started, run `litedown::roam()` and it will launch a file browser to let
you preview everything that can be rendered to HTML, such as `.md`, `.Rmd`, and
`.R` files (@sec:live-preview).

The rendering takes place in memory, which means it will not render `.html`
files on your disk, unless you request so. The page will be automatically
updated as you edit and save a file. This update-on-save feature can be turned
off, and then you can manually refresh the page to re-render the file whenever
you want.

You can open any file in your editor or the default system application by
clicking a button in the browser. You can also render a file or a project in a
new R session by clicking a button in the browser.

### Precise parser

The R Markdown parser stores the precise line and column numbers of code
elements. The location information is used in various places. For example, when
an error occurs, you will get a message that tells you the precise location in
the source. In editors that support ANSI links (such as the RStudio IDE), you
can even click on the message to go to a specific line/column in the source
document, so you can quickly and easily know where the error occurred exactly.
When previewing `.Rmd` documents with `roam()`, you will see line numbers on the
left side of code blocks. Clicking on these numbers will bring you to the lines
in the source.

Due to the fact that the parser is based on **commonmark** (instead of solely on
regular expressions like **knitr**'s parser), it can precisely recognize code
chunks and inline code, which means code within other verbatim code blocks or
comments will be untouched. For example:

````` md
````md
Below is not a code chunk but verbatim content
inside a fenced code block (with four backticks).

```{r}
1 + 1
```

Inline code expressions like `{r} 1+1` are not
parsed, either.
````
`````

```` html
<!--
Feel free to comment out anything, such as a code chunk:

```{r}
1 + 1
```
-->
````

### Versioned CSS/JS assets

By default, **litedown** produces bare minimal HTML output, but many features
can be enabled by adding CSS/JS assets (@chp:assets). You can freely choose
whatever versions that work for you to avoid potential future breakage when the
assets are upgraded. The assets are not bundled in the **litedown** package, but
hosted on CDN, so updating **litedown** will not update your assets.

The CSS and JS code for commonly used features do not depend on any frameworks
such as Bootstrap or jQuery. They are simply vanilla code written from scratch.
No generator was used (such as SCSS). The code is often relatively short, so you
could just fork the code, modify it, publish your own version, and use it if you
are not satisfied with the original version.

### Chunk options management

Chunk options are managed by an environment, i.e., `litedown::reactor()`
(@sec:reactor). Using an environment (as compared to a list like **knitr**'s
`opts_chunk`) means you can access the up-to-date chunk options anywhere,
because environments in R are mutable. I cannot explain how awkward **knitr**'s
`opts_current` has been. It is basically a lie—chunk options that you get from
`opts_current` are not necessarily "current".

### Non-linear order of execution

A document does not have to be compiled in the linear order. With the chunk
option `order`, you can specify code chunks and inline code expressions to be
executed in an arbitrary order (@sec:option-order). For example, if you have an
abstract in the beginning of a document, and the abstract needs to use a number
calculated at the end of the document, you can let the abstract be compiled in
the end, although it appears earlier in the document.

I guess some users may want to kill me upon knowing this feature, and some may
send me flowers (although I'm not sure if they want to thank me or prepare the
flowers for my funeral). For those who want to kill me, please note that this
feature does not mean **litedown** is as awful as Jupyter notebooks. It means
you can specify a *fixed* order to execute code in the document. The order does
not have to be from beginning to end, but it is *deterministic*. In other words,
it does not mean you run code chunks in an arbitrary or random order that can
detriment reproducibility.

### Time code chunks

If you want to figure out which code chunks are time-consuming, simply set the
chunk option `litedown::reactor(time = TRUE)` (@sec:option-time) in the
beginning, and put `litedown::timing_data()` in the last code chunk. It will
tell you detailed information. In the `roam()` preview, the data will also
contain links to specific lines of code chunks, so you click to jump to a
specific code chunk.

### Table output for data frames by default

Rectangular objects such as data frames (including tibbles) and matrices are
printed as tables by default (@sec:option-tab), and the number of rows is
limited to 10 by default to avoid generating huge tables by accident.

### Relieved pain of paths

File paths (such as image paths) have been a mess in **knitr**. My deepest
apologies for that. I have worked much harder in **litedown** in this regard.

### Output in memory or to disk

Functions such as `mark()`, `fuse()`, and `fuse_book()` can operate in memory
without writing to disk. By default, if you pass a file input, you get a file
output; if you pass text input, you get text output.

### A new cache system

You can feel more confident with using the chunk option `cache = TRUE` in
**litedown** than in **knitr**. The new cache system is more robust and
intelligent (@sec:option-cache).

### A book is a single HTML file

Unlike **bookdown** and Quarto, `litedown::fuse_book()` renders multiple input
files to a single output file. Yes, your whole precious book is in a single
(HTML or PDF) file. Grab and go.

The assumption of single-file output for books has made several things a lot
easier. For example, if you want to search within the book, just press
`Ctrl + F` or `Command + F` in your browser as you usually do on any other web
page. You do not need a client-side search library. It is also quicker to jump
between chapters because they are on the same page. If you want to print the
HTML version of the book to PDF, just press `Ctrl + P` or `Command + P`.

I know you have a concern: wouldn't the single HTML file be too heavy for the
browser? The answer is: you should be fine if you do not have too many images.
If you do, do not base64 encode them (which is the default), and you can also
lazy-load images to make the book load faster.

### R scripts as first-class citizens

An R script has the same status as an R Markdown document (@sec:r-scripts). You
can write Markdown in `#'` comments, and chunk options in `#|` comments (both
are optional). These are the only two rules to remember. Compared to
`knitr::spin()`, the rules are much simpler. Besides, R scripts are also
carefully parsed into "code chunks", so their line numbers work as well as R
Markdown's line numbers.

Any application that you can create with R Markdown can be created with R
scripts, such as books and websites.

### Clean HTML output

The HTML output generated from **litedown** is very clean. For example, code
blocks in HTML output contain plain code, instead of full of `<span>` tags with
random attributes. Clean HTML output means the output file size is smaller, and
more importantly, it is easier to inspect the differences between two versions
of output (e.g., in Git). Every time when you update the source document, you
can know more clearly what has changed in the output, which can help you avoid
unexpected changes before publishing the output.
